\vspace{-0.25em}\thetitle

### 2009-07-28 paper.tex
...

This work begins with a particular way of representing flow behaviors as vectors.
% and the behaviors of entire traffic traces as large, sparse matrices.
% This approach was first proposed in by Karpinski~\emph{et~al.}~\cite{Karpinski08}, who called this general type of representation \emph{linear representations}.
The representation is quite simple.
For each feature of a flow, we represent that aspect of the flow's behavior as a \emph{feature-frequency vector}:
a vector having a dimension for each possible value of the feature and whose coordinates are the relative frequency of values.
For example, the vector representing the distribution of packet sizes of a flow having four 40-byte and two 145-bytes packets is
\begin{align}
  \frac{1}{6}\parens{4\vec{e}_{40} + 2\vec{e}_{145}}.
\end{align}
% The dimension of this representation is 1500 because that is typically the maximum transfer unit (\caps{MTU}) of local networks.
Different aspects of flow behavior can be represented in this way, and these representations can be combined by taking the direct sum of their representation vectors.

@@

A linear representation, as proposed by Karpinski~\emph{et~al.}~\cite{Karpinski08}, maps sets of flows into a vector space such that disjoint union of sets becomes addition in the vector space:
\begin{align}
  \Phi(s_1 \disunion s_2) = \Phi(s_1) + \Phi(s_2).
\end{align}
The representation function should map each set of flows to a vector that in some sense describes their collective behavior.
For example: let $\Phi(s) \in \R^1$ be the number of packets in the flows in $s$.
Since the number of packets in the disjoint union of sets of flows is the sum of the number in each set individually, this representation is linear.
Given two representations, their direct sum is readily seen to also be a linear representation:
\begin{align}
  \Phi(s) = \Phi_1(s) \directsum \Phi_2(s).
\end{align}

A particularly powerful class of linear representations describes the behavior of collections of flows using histogram vectors to count the frequencies of various features.
A 1500-dimensional vector can be used to count the frequency of packet sizes seen, from one byte up to the typical \caps{MTU} of 1500 bytes.
In this representation, $\Phi(s)$ has as its $i$th entry the number of packets in $s$ of size $i$.
Since \caps{IP} protocols range from 0 to 255, the frequency of types in a set of flows can be represented in 256 dimensions.
Histogram representations for port numbers and inter-packet intervals are detailed in \Section{methodology}.
% Source and destination port numbers can similarly be represented using 65536-dimensional vectors.
% For practical purposes, however, this is too large a vector space, so we represent port numbers in 1000-dimensions, and reduce port numbers modulo 1000.
% Finally, we represent inter-packet intervals by first quantizing into 1500 bins of interval values as described by Karpinski~\emph{et~al.} and counting the number of intervals in flows occurring in each bin.
The representation we use is a direct sum of these histogram representations.
% The combination is a representation in 5256 dimensions.
An entire trace is represented as a matrix of such representation vectors---one flow per row.
This paper explores the linear structure of such matrices.

@@

To see the linear structure of these representation matrices, we apply one of the most powerful tools of linear algebra:
the singular value decomposition (\caps{SVD}).
This matrix decomposition can be applied to any real matrix, factoring it into a product of three real matrices:
\begin{align}
  X = US\trans{V}.
\end{align}
$U$ and $V$ are orthogonal and $S$ is diagonal, with entries of nonincreasing nonnegative value.
Among the useful properties of this factorization, is that the first $k$ columns of $US$ are an optimal $k$-dimensional reduced model of $X$.\svdnote
This property allows us to visualize the structure of $X$ using initial columns of $US$.
However, we are interested only interested in direction since
vectors differing only by magnitude lie in the same one-dimensional subspace.
To visualize direction while eliminating distance, we project all data points onto the unit-sum hyperplane from the origin.\projectionnote

\Figure{svd} illustrates the results of applying \caps{SVD} dimension reduction to sample matrices of six different traffic traces.
The figures show scatter plots of the most significant two projected \caps{SVD} coordinates for a matrix of 5000 sample flows from each trace.
% The original data encodes packet size distributions, inter-packet interval distributions, \caps{IP} protocol numbers, and port numbers for a sample of 5000 flows from each trace.
% The exact details of how the linear representation encodes these features will be discussed in \Section{methodology}.
Linear structure is visible in all six:
flow behaviors are heavily concentrated near low-dimensional subspaces.
Behaviors concentrated along a one-dimensional subspace appear as dark ``spots'' where many points have been plotted;
behaviors near two-dimensional subspaces appear as lines;
behaviors on higher-dimensional subspaces appear as ``smears.''
% There is a smattering of outliers, not especially near such structures.

The presence of linear structure in the matrix representations of several independent traffic traces immediately raises two significant questions.
The existential question: \emph{Why does this linear structure exist?}
Followed, of course, by the practical question: \emph{How can we exploit this structure?}
Neither question can be answered fully in a single paper, but we will begin to address the existential problem in the next section.
In the rest of the paper we apply this insight to the eminently practical problem of predicting the behavior of individual flows from the observation of only a few of their initial packets.

>>>>

This work begins with a particular way of representing flow behaviors as vectors.
% and the behaviors of entire traffic traces as large, sparse matrices.
% This approach was first proposed in by Karpinski~\emph{et~al.}~\cite{Karpinski08}, who called this general type of representation \emph{linear representations}.
The representation is quite simple.
For each feature of a flow, we represent that aspect of the flow's behavior as a \emph{feature-frequency vector}:
a vector having a dimension for each possible value of the feature and whose coordinates are the relative frequency of values.
For example, the vector representing the distribution of packet sizes of a flow having four 40-byte and two 145-bytes packets is
\begin{align}
  \text{size} = \frac{1}{6}\parens{4\vec{e}_{40} + 2\vec{e}_{145}}.
\end{align}
% The dimension of this representation is 1500 because that is typically the maximum transfer unit (\caps{MTU}) of local networks.
Different aspects of flow behavior can be represented in this way, and these representations can be combined by taking the direct sum of their representation vectors.

@@

\newfootnote{\directsumnote}{The direct sum is the vector space analogue of a cross product of sets.}

Before we can present our motivating observation about the linear structure of network traffic, we must explain how flow behaviors are represented as feature-frequency vectors. spaces.
The low-level details of this representation are given in \Section{representation}.
For now, what the reader must know is that each feature is represented by a vector where each possible value of the feature correspond to a dimension and the normalized frequency of that value occurring is the corresponding coordinate in the vector.
We have already given the example of packet size distribution in the introduction.
The vectors are normalized in that the sum of the coordinates of each feature-frequency vector is unity.
To combine multiple feature-frequency vectors into a composite description of the overall behavior of the flow, we take the direct sum their representations:\directsumnote
\begin{align}
  \text{flow} =
  \text{size} \directsum
  \text{ival} \directsum
  \text{type} \directsum
  \text{ports} \directsum
  \text{pkts}.
\end{align}
The behavior of a feature across a collection of flows can be represented as a matrix where each row represents a flow:
\begin{align}
  \text{Size} = \begin{bmatrix}
    \text{size}_1 \\
    \vdots \\
    \text{size}_m
  \end{bmatrix}.
\end{align}
The overall behavior of the collection of flows then becomes a concatenation of these feature matrices:
\begin{align}
  X = [ ~
  \text{Size} ~
  \text{Ival} ~
  \text{Type} ~
  \text{Ports} ~
  \text{Pkts} ~
  ].
\end{align}
It is this matrix, representing the total behavior of a collection of flows, to which we apply our analysis.

@@

To see the linear structure of these representation matrices, we apply one of the most fundamental tools of linear algebra:
the singular value decomposition (\caps{SVD}).
This matrix decomposition can be applied to any real matrix, factoring it into a product of three real matrices:
\begin{align}
  X = US\trans{V}.
\end{align}
$U$ and $V$ are orthogonal and $S$ is diagonal, with entries of nonincreasing nonnegative value.
Among the useful properties of this factorization, is that the first $k$ columns of $US$ are an optimal $k$-dimensional reduced model of $X$.\svdnote
This property allows us to visualize the structure of $X$ using initial columns of $US$---and this is precisely what we do.
However, it must be done in a particular way.
First determine an appropriate value for $k$.
Here we take the number of dimensions necessary to account for 75\% of the variance seen in $X$;
other approaches work well too.
Discarding the rest of the dimensions has the effect of ``denoising'' the data.
The vectors of the remaining $k$-dimensional reduction of $X$ may have different magnitudes, but we are only interested in direction, since vectors differing only be magnitude lie in the same one-dimensional subspace.
Therefore, we project each row vector through the origin onto the unit-sum hyperplane.\projectionnote
It is the coordinates of this reduced, projected data which exhibit clear linear structure.

\Figure{svd} shows scatter plots of most significant two coordinates when this procedure is applied to a random sample of 5000 flows from each of six unrelated traffic traces.
Clear linear structure is visible in all six traces:
flow behaviors are heavily concentrated near low-dimensional subspaces.
Behaviors concentrated along a one-dimensional subspace appear as dark ``spots'' where many points have been plotted;
behaviors near two-dimensional subspaces appear as lines;
behaviors on higher-dimensional subspaces appear as ``smears.''

The presence of linear structure in the matrix representations of several independent traffic traces immediately raises two significant questions.
The existential question: \emph{Why does this linear structure exist?}
Followed, of course, by the practical question: \emph{How can we exploit this structure?}
Neither question can be answered fully in a single paper, but we will begin to address the existential problem in the next section.
In the rest of this paper we apply this insight to the eminently practical problem of predicting the behavior of individual flows from the observation of only a few of their initial packets.
